---
title: "cran-comments"
output: github_document
---

## Resubmission 2

In this version I have

* changed the title to CRAN-suggested title case (I gave up..);
* explained the GSED acronym;
* removed "Tools for .." in Description field.


## Resubmission

This is a resubmission. In this version I have

* Rewritten the `Description:` field according to the suggestions. Thanks.
* Removed the `LICENSE` and `LICENSE.md` files from the package root, and added a `Copyright:` field to the DESCRIPTION file. My impression was that I needed to the `LICENSE.md` to make it valid GPL-3, but I am happy to live without it.

```
The Title field should be in title case. Current version is:
'D-score for Child Development'
In title case that is:
'D-Score for Child Development'
```

I did not alter the package title. The submitted title `D-score for Child Development` *is* title case. The expression `D-score` is a single word, like `F-test`, `P-value`, `K-means` or `Q-technique`, all spelled according to the Cambridge Dictionary of Statistics. I believe the correct spelling is `D-score`, not `D-Score`. I have consistently used `D-score` in all publications on the topic, so changing the word to `D-Score` would create confusion. My suggestion is to add `D-score` to the exceptions list of the function that scans for title case.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Reason 

This is a new release.

## Checks

Package built by

```{r eval = FALSE}
library("devtools")
build(manual = TRUE)
```

```{bash eval=FALSE}
R CMD CHECK /Users/buurensv/Package/dscore/dscore_1.0.0.tar.gz

...
* checking data for non-ASCII characters ... NOTE
  Note: found 3 marked UTF-8 strings
...

Status: 1 NOTE
```

As I believe UTF-8 codes are supported by CRAN, we can ignore the note.

## Test environments
* local OS X install, 10.15.2, R 3.6.2
* win-builder, using `devtools::check_win_devel()`
* rhub

`devtools::check_win_devel()` resulted in 1 NOTE:

```
* checking CRAN incoming feasibility ... NOTE
Maintainer: 'Stef van Buuren <stef.vanbuuren@tno.nl>'

New submission

License components with restrictions and base license permitting such:
  GPL-3 + file LICENSE
File 'LICENSE':
  YEAR: 2020
  COPYRIGHT HOLDER: Stef van Buuren; Iris Eekhout; Arjan Huizing

Possibly mis-spelled words in DESCRIPTION:
  DAZ (11:25)
  GSED (13:5)
  dscore (8:18)

The Title field should be in title case. Current version is:
'D-score for Child Development'
In title case that is:
'D-Score for Child Development'
```

I believe this is standard for a new package.
The title should read "D-score".

```{r eval=FALSE}
check_rhub()
```


This throws the following:

```
Windows Server 2008 R2 SP1, R-devel, 32/64 bit

dscore 1.0.0: NOTE

* checking CRAN incoming feasibility ... NOTE
Maintainer: 'Stef van Buuren <stef.vanbuuren@tno.nl>'

New submission

License components with restrictions and base license permitting such:
  GPL-3 + file LICENSE
File 'LICENSE':
  YEAR: 2020
  COPYRIGHT HOLDER: Stef van Buuren; Iris Eekhout; Arjan Huizing

Possibly mis-spelled words in DESCRIPTION:
  DAZ (11:25)
  GSED (13:5)
  dscore (8:18)

The Title field should be in title case. Current version is:
'D-score for Child Development'
In title case that is:
'D-Score for Child Development'

* checking for non-standard things in the check directory ... NOTE
Found the following files/directories:
  'dscore-Ex_i386.Rout' 'dscore-Ex_x64.Rout' 'examples_i386'
  'examples_x64' 'tests_i386' 'tests_x64'
  
0 errors ✓ | 0 warnings ✓ | 2 notes x

------------

Ubuntu Linux 16.04 LTS, R-release, GCC

dscore 1.0.0: NOTE

* checking CRAN incoming feasibility ... NOTE
Maintainer: ‘Stef van Buuren <stef.vanbuuren@tno.nl>’

New submission

License components with restrictions and base license permitting such:
  GPL-3 + file LICENSE
File 'LICENSE':
  YEAR: 2020
  COPYRIGHT HOLDER: Stef van Buuren; Iris Eekhout; Arjan Huizing

Possibly mis-spelled words in DESCRIPTION:
  DAZ (11:25)
  dscore (8:18)
  GSED (13:5)

The Title field should be in title case. Current version is:
‘D-score for Child Development’
In title case that is:
‘D-Score for Child Development’

------------
  
Fedora Linux, R-devel, clang, gfortran

dscore 1.0.0: NOTE

* checking CRAN incoming feasibility ...NB: need Internet access to use CRAN incoming checks
 NOTE
Maintainer: ‘Stef van Buuren <stef.vanbuuren@tno.nl>’

License components with restrictions and base license permitting such:
  GPL-3 + file LICENSE
File 'LICENSE':
  YEAR: 2020
  COPYRIGHT HOLDER: Stef van Buuren; Iris Eekhout; Arjan Huizing

Possibly mis-spelled words in DESCRIPTION:
  DAZ (11:25)
  GSED (13:5)
  dscore (8:18)

The Title field should be in title case. Current version is:
‘D-score for Child Development’
In title case that is:
‘D-Score for Child Development’

------------
  
Debian Linux, R-devel, GCC ASAN/UBSAN

dscore 1.0.0: PREPERROR

Status for build dscore_1.0.0.tar.gz-34f374f542484e50afb89ddbe9988cbd
Status: success 
Duration: 1 hour 7 minutes 6.9 seconds

...
11212#> Build step 'Send files or execute commands over SSH' changed build result to SUCCESS
11213#> Pinging https://builder.r-hub.io/build/SUCCESS/dscore_1.0.0.tar.gz-34f374f542484e50afb89ddbe9988cbd/2019-12-30T16:42:30Z
11214#> {"status":"ok"}
11215#> Finished: SUCCESS
```

I do not understand all details of these checks. To me, it appears that 
the message are unrelated to the `dscore` package.

## Downstream dependencies

There are no downstream dependencies for this package.
