---
title: "cran-comments"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Reason 

This is a new release.

## Checks

Package built by

```{r eval = FALSE}
library("devtools")
build(manual = TRUE)
```

```{bash eval=FALSE}
R CMD CHECK /Users/buurensv/Package/dscore/dscore_1.0.0.tar.gz

...
* checking data for non-ASCII characters ... NOTE
  Note: found 3 marked UTF-8 strings
...

Status: 1 NOTE
```

As I believe UTF-8 codes are supported by CRAN, we could probably ignore the note.

## Test environments

* local OS X install, 10.15.2, R 3.6.2
* win-builder, using `devtools::check_win_devel()`

Status:
`devtools::check_win_devel()` resulted in:

```
* checking CRAN incoming feasibility ... NOTE
Maintainer: 'Stef van Buuren <stef.vanbuuren@tno.nl>'

New submission

License components with restrictions and base license permitting such:
  GPL-3 + file LICENSE
File 'LICENSE':
  YEAR: 2020
  COPYRIGHT HOLDER: Stef van Buuren; Iris Eekhout; Arjan Huizing

Possibly mis-spelled words in DESCRIPTION:
  DAZ (11:25)
  GSED (13:5)
  dscore (8:18)

The Title field should be in title case. Current version is:
'D-score for Child Development'
In title case that is:
'D-Score for Child Development'
```

I believe these are all beninge notes that can be ignored.

```{r eval=FALSE}
check_rhub()
```


This throws various warnings and notes:

```{r eval=FALSE}
Ubuntu Linux 16.04 LTS, R-release, GCC
Maintainer: ‘Stef van Buuren <stef.vanbuuren@tno.nl>’

New submission

License components with restrictions and base license permitting such:
  GPL-3 + file LICENSE
File 'LICENSE':
  YEAR: 2020
  COPYRIGHT HOLDER: Stef van Buuren; Iris Eekhout; Arjan Huizing

Possibly mis-spelled words in DESCRIPTION:
  DAZ (11:25)
  dscore (8:18)
  GSED (13:5)

The Title field should be in title case. Current version is:
‘D-score for Child Development’
In title case that is:
‘D-Score for Child Development’
* checking Rd cross-references ... NOTE
Package unavailable to check Rd xrefs: ‘sirt’


---
  
Fedora Linux, R-devel, clang, gfortran

ERRORS:
* checking package dependencies ... ERROR
Package suggested but not available: ‘kableExtra’

The suggested packages are required for a complete check.
Checking can be attempted without them by setting the environment
variable _R_CHECK_FORCE_SUGGESTS_ to a false value.

See section ‘The DESCRIPTION file’ in the ‘Writing R Extensions’
manual.
NOTES:
* checking CRAN incoming feasibility ...NB: need Internet access to use CRAN incoming checks
 NOTE
Maintainer: ‘Stef van Buuren <stef.vanbuuren@tno.nl>’

License components with restrictions and base license permitting such:
  GPL-3 + file LICENSE
File 'LICENSE':
  YEAR: 2020
  COPYRIGHT HOLDER: Stef van Buuren; Iris Eekhout; Arjan Huizing

Possibly mis-spelled words in DESCRIPTION:
  DAZ (11:25)
  GSED (13:5)
  dscore (8:18)

The Title field should be in title case. Current version is:
‘D-score for Child Development’
In title case that is:
‘D-Score for Child Development’

----
  
dscore 1.0.0: PREPERROR
Build ID:	dscore_1.0.0.tar.gz-37d55a48841c4c30be2b59d8eb695f5a
Platform:	Debian Linux, R-devel, GCC ASAN/UBSAN
Submitted:	1 hour 22 minutes 15 seconds ago
Build time:	1 hour 21 minutes 57.8 seconds


Status for build dscore_1.0.0.tar.gz-37d55a48841c4c30be2b59d8eb695f5a
Status: success 
Duration: 1 hour 22 minutes 0.2 seconds
...
Pinging https://builder.r-hub.io/build/SUCCESS/dscore_1.0.0.tar.gz-37d55a48841c4c30be2b59d8eb695f5a/2019-12-30T13:23:16Z
{"status":"ok"}
Finished: SUCCESS
```


I do not understand all details of these checks. To me, it appears that 
the message are unrelated to the dscore package.

## Downstream dependencies

There are no downstream dependencies for this package.

