---
title: "cran-comments"
output: github_document
---

## Reason 

Release 1.1.0 contains a few bug fixes.

## Test environments

* local OS X install, 10.15.3, R 3.6.2
* win-builder, using `devtools::check_win_devel()`
* rhub

### Local build

Package built by

```{r eval = FALSE}
library("devtools")
build(manual = TRUE)
```

```{bash eval=FALSE}
R CMD CHECK /Users/buurensv/Package/dscore/dscore_1.1.0.tar.gz

...
* checking data for non-ASCII characters ... NOTE
  Note: found 3 marked UTF-8 strings
...

Status: 1 NOTE
```

UTF-8 codes are supported by CRAN, so I assume we can ignore the note.

### WIN_DEVEL

`devtools::check_win_devel()` resulted in:

```
Status: OK
```

### RHUB

```{r eval=FALSE}
check_rhub()
```

The result is:

```
Platform:	Windows Server 2008 R2 SP1, R-devel, 32/64 bit
Status: 1 NOTE

Found the following files/directories:
  'dscore-Ex_i386.Rout' 'dscore-Ex_x64.Rout' 'examples_i386'
  'examples_x64' 'tests_i386' 'tests_x64'
* checking for detritus in the temp directory ... OK
* DONE
```

```
Platform:	Ubuntu Linux 16.04 LTS, R-release, GCC
Status: OK
```

```
Platform:	Fedora Linux, R-devel, clang, gfortran
Status: OK
```

```
Platform:	Debian Linux, R-devel, GCC ASAN/UBSAN
dscore 1.1.0: PREPERROR

SSH: Transferred 93 file(s)
Build step 'Send files or execute commands over SSH' changed build result to SUCCESS
Pinging https://builder.r-hub.io/build/SUCCESS/dscore_1.1.0.tar.gz-9cc3cabcaec042f3945b1f48cc7c0044/2020-02-23T15:31:45Z
{"status":"ok"}
Finished: SUCCESS
```

I believe these are benign messages, unrelated to the `dscore` package.

## Downstream dependencies

There are no downstream dependencies for this package.
