---
title: "cran-comments"
output: github_document
---

## Reason for update

`dscore 1.4.0` extends the itembank and resolves some problems

## Test environments

* local OS X install, 11.0.1, R 4.0.3
* win-builder, using `devtools::check_win_devel()`
* rhub

### Local build

Package built by

```{r eval = FALSE}
library("devtools")
build(manual = TRUE)
```

```{bash eval=FALSE}
R CMD CHECK /Users/buurensv/Package/dscore/dscore_1.4.0.tar.gz

Status: OK
```

### WIN_DEVEL

`devtools::check_win_devel()` resulted in:

```
Status: OK
```

### RHUB

```{r eval=FALSE}
check_rhub()
```

The result is:

```
── dscore 1.4.0: ERROR

Build ID:	dscore_1.4.0.tar.gz-97d617cf0f4d458aae592e265a2f8872
Platform:	Windows Server 2008 R2 SP1, R-devel, 32/64 bit
Submitted:	6 minutes 59.9 seconds ago
Build time:	6 minutes 57.7 seconds

* checking examples ...
** running examples for arch 'i386' ... ERROR
Running examples in 'dscore-Ex.R' failed
The error most likely occurred in:

> ### Name: get_age_equivalent
> base::assign(".ptime", proc.time(), pos = "CheckExEnv")
> ### Aliases: get_age_equivalent
> 
Error in loadNamespace(x) : there is no package called 'utf8'
```

SvB: The problem with the `utf8` package is outside my reach. 

```
dscore 1.4.0: OK
Build ID:	dscore_1.4.0.tar.gz-b7c485c90af2483588faae90192af7d9
Platform:	Ubuntu Linux 16.04 LTS, R-release, GCC
Submitted:	47 minutes 33.8 seconds ago
Build time:	47 minutes 31.6 seconds
```

```
dscore 1.4.0: OK
Build ID:	dscore_1.4.0.tar.gz-07cb0e7218cb4fc2920b459722c0dd79
Platform:	Fedora Linux, R-devel, clang, gfortran
Submitted:	54 minutes 40.1 seconds ago
Build time:	54 minutes 38.3 seconds
```

```
dscore 1.4.0: PREPERROR
Build ID:	dscore_1.4.0.tar.gz-bd4ae33d5a6b4b8caaff07a0d493a465
Platform:	Debian Linux, R-devel, GCC ASAN/UBSAN
Submitted:	1 hour 14 minutes 53.8 seconds ago
Build time:	1 hour 14 minutes 42.8 seconds

> ERROR: dependencies ‘xml2’, ‘rvest’ are not available for package ‘kableExtra’
> ERROR: dependencies ‘CDM’, ‘TAM’, ‘pbv’ are not available for package ‘sirt’```
```

SvB: Test environment Debian Linux is incomplete. Unlikely to be related to the `dscore` package. 


## Downstream dependencies

There are no downstream dependencies for this package.
