---
title: "cran-comments"
output: github_document
editor_options: 
  chunk_output_type: console
---

## Reason for update

`dscore 1.9.0` is a major update providing new features and resolving some problems

## Resubmission 1

`rhub_setup()` did not add `^.github$` to `.Rbuildignore`. Now done manually.

## Resubmission 2

Removed markup from DOI field in CITATION file


## Test environments

### Local

```{r}
R.Version()
```

## Local check

Package built by

```{r eval = FALSE}
library("devtools")
build()
```

```{bash eval=FALSE}
R CMD CHECK dscore_1.9.0.tar.gz
```

Status: OK

## win-builder

```{r eval=FALSE}
devtools::check_win_devel()
```

### WIN_DEVEL

`devtools::check_win_devel()` resulted in:

```
Found the following (possibly) invalid URLs:
  URL: https://tnochildhealthstatistics.shinyapps.io/dcalculator
    From: inst/doc/scoring_GSED.html
    Status: 202
    Message: Accepted
  URL: https://tnochildhealthstatistics.shinyapps.io/dcalculator/
    From: README.md
    Status: 202
    Message: Accepted

Found the following (possibly) invalid DOIs:
  DOI: https://doi.org/10.1201/9781003216315
    From: inst/CITATION
    Status: 404
    Message: Not Found
```

These URLs are reachable by the browser, so I assume these are false positives.

### RHUB

```{r eval=FALSE}
rhub::rhub_check()
```

Using five builds: Results <https://github.com/D-score/dscore/actions> are OK.

## Downstream dependencies

There are no downstream dependencies for this package.
