---
title: "cran-comments"
output: github_document
editor_options: 
  chunk_output_type: console
---

## Reason for update

`dscore 1.8.0` provides new features and resolves some problems

## Test environments

### Local

```{r}
R.Version()
```

## Local check

Package built by

```{r eval = FALSE}
library("devtools")
build()
```

```{bash eval=FALSE}
R CMD CHECK dscore_1.8.0.tar.gz
```

Status: OK

## win-builder

```{r eval=FALSE}
devtools::check_win_devel()
```

### WIN_DEVEL

`devtools::check_win_devel()` resulted in:

```
* checking CRAN incoming feasibility ... [11s] NOTE
Maintainer: 'Stef van Buuren <stef.vanbuuren@tno.nl>'

Found the following (possibly) invalid URLs:
  URL: https://support.posit.co/hc/en-us/articles/201141096-Getting-Started-with-R
    From: inst/doc/scoring_GSED.html
    Status: 403
    Message: Forbidden
```

The URL is reachable by browser. I assume this status results from a setting made by Posit.

### RHUB

```{r eval=FALSE}
check_rhub()
```

The result is: `SUCCESS` for all four builds


## Downstream dependencies

There are no downstream dependencies for this package.
