---
output: github_document
always_allow_html: yes
bibliography: [references.bib]
biblio-style: apalike
link-citations: yes
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
options(width = 130)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# dscore

<!-- badges: start -->
[![Lifecycle: maturing](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://www.tidyverse.org/lifecycle/#maturing)
[![](https://img.shields.io/badge/github%20version-0.38-orange.svg)](https://github.com/stefvanbuuren/dscore/rename)
<!-- badges: end -->

The `dscore` package contains tools to

- Calculate D-score from item level responses
- Transform the D-scores into DAZ, age-standardised Z-scores
- Map your item names to the GSED convention

The required input consists of *item level* responses on milestones
from widely used instruments for measuring child development.

## Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("remotes")
remotes::install_github("stefvanbuuren/dscore@rename")
```

## Example

### Inspect the data

The `milestones` dataset in the `dscore` package contain responses
of 27 preterm children measured at various age between birth and 2.5 
years on the Dutch Development Instrument (DDI). The dataset looks 
like:

```{r milestones}
library(dscore)
head(milestones[, c(1, 3, 4, 9:15)])
```

Each row corresponds to a visit. Most children have three or four visits.
Columns starting with `ddi` hold the responses on DDI-items. A 
`1` means a PASS, a `0` means a FAIL, and `NA` means that the item was 
not administered.

The `get_labels()` function obtains the labels of the milestones, e.g.,

```{r}
items <- names(milestones)[9:15]
labels <- get_labels(items)
print(data.frame(items, labels), right = FALSE, row.names = FALSE)
```

### Calculate the D-score and DAZ

The `milestones` dataset has properly named columns that identifies each
item. Calculating the D-score and DAZ is then done by:

```{r}
ds <- dscore(milestones)
dim(ds)
```

where `ds` is a `data.frame` with the same number of rows than the input 
data. The first six rows are

```{r}
head(ds)
```

The table below provides the interpretation of each column: 

Name | Interpretation
---- | -------------
a    | Decimal age
n    | Number of items used to calculate D-score
p    | Percentage of passed milestones
d    | D-score estimate, mean of posterior
sem  | Standard error of measurement, standard deviation of the posterior
daz  | D-score corrected for age

### Graphs of D-score and DAZ

The `milestones` data and the result can be combined as

```{r}
md <- cbind(milestones, ds)
```

The individual developmental curves of 27 children can be plotted.

```{r graphD, dpi = 144}
library(ggplot2)
ggplot(md, aes(x = a, y = d, group = id, color = sex)) + 
  xlab("Age (in years)") + 
  ylab("D-score") +
  geom_line(lwd = 0.1) +
  geom_point(size = 1) +
  scale_colour_hue(l = 45) +
  theme_light()
```

The DAZ is an age-standardized D-score with a standard normal distribution
with mean 0 and variance 1. The individual DAZ curves are plotted as

```{r graphDAZ, dpi = 144}
ggplot(md, aes(x = a, y = daz, group = id, color = sex)) + 
  theme_light() +
  geom_rect(xmin = 0, xmax = 3, ymin = -2, ymax = 2, 
            fill = "grey97", linetype = 0, show.legend = FALSE) +
  geom_hline(yintercept = -1:1, colour = "white", lwd = 0.5) +
  coord_cartesian(ylim = c(-4, 4)) +
  geom_line(lwd = 0.1) +
  geom_point(size = 1) +
  scale_colour_hue(l = 45) +
  xlab("Age (in years)") +
  ylab("DAZ")
```

Note that the D-scores and DAZ are a little lower than average. The 
explanation here is that these all children are born preterm.

### Mapping items names

The `dscore()` function assumes that the item names in your data follow 
the GSED naming convention. This convention is a nine-position word that 
identifies the instrument, the domain, the administration mode and the 
item number. You may decompose an item names into these components, 
as follows:

```{r}
decompose_itemnames(items)
```

which gives four components of the seven items. 

In practice, you will need to spend some time in renaming your own 
item names according to the GSED convention. In order to ease this 
process, the `dscore` package offers several functions that can help.

First of all, the measurement instrument in your data needs to be one 
of the instruments supported by the `dscore` package. Here's the 
table of instrument names and number of items that are currently 
defined in the package:

| Code   | Instrument                                              | Items | dutch |  gcdg |  gsed |
| ------ | ------------------------------------------------------- | -----:|------:|------:|------:|
| `aqi`  | Ages & Stages Questionnaires-3                          |  230  |       |   29  |   17  |
| `bar`  | Barrera Moncada                                         |   22  |       |   15  |   13  |
| `bat`  | Battelle Development Inventory and Screener-2           |  137  |       |       |       |
| `by1`  | Bayley Scales for Infant and Toddler Development-1      |  156  |       |   85  |   76  |
| `by2`  | Bayley Scales for Infant and Toddler Development-2      |  121  |       |   16  |   16  |
| `by3`  | Bayley Scales for Infant and Toddler Development-3      |  320  |       |  105  |   67  |
| `cro`  | Caregiver Reported Early Development Instrument (CREDI) |  149  |       |       |   62  |
| `ddi`  | Dutch Development Instrument (Van Wiechenschema)        |   77  |   76  |   65  |   64  |
| `den`  | Denver-2                                                |  111  |       |   67  |   50  |
| `dmc`  | Developmental Milestones Checklist                      |   66  |       |       |   43  |
| `gri`  | Griffiths Mental Development Scales                     |  312  |       |  104  |   93  |
| `iyo`  | Infant and Young Child Development (IYCD)               |   90  |       |       |   55  |
| `kdi`  | Kilifi Developmental Inventory                          |   69  |       |       |       |
| `mac`  | MacArthur Communicative Development Inventory           |    6  |       |    3  |    3  |
| `mds`  | WHO Motor Development Milestones                        |    6  |       |       |    1  |
| `mdt`  | Malawi Developmental Assessment Tool (MDAT)             |  136  |       |       |  126  |
| `peg`  | Pegboard                                                |    2  |       |    1  |    1  |
| `pri`  | Project on Child Development Indicators (PRIDI)         |   63  |       |       |       |
| `sbi`  | Stanford Binet Intelligence Scales-4/5                  |   33  |       |    6  |    1  |
| `sgr`  | Griffiths for South Africa                              |   58  |       |   19  |   19  |
| `tep`  | Test de Desarrollo Psicomotor (TEPSI)                   |   61  |       |   33  |   31  |
| `vin`  | Vineland Social Maturity Scale                          |   50  |       |   17  |   17  |
| `hyp`  | [special codes, hypothetical instrument]                |    5  |       |       |       |
|        |                                                         | 2280  |   76  |  565  |  807  |

If your instrument is not here, you cannot calculate the D-score. But 
even if your instrument is in the table, there is no garantee that it
can be used for the D-score. Instruments are tied to the D-score by 
means of a *measurement model*. The measurement model provides the
*key* for converting the item responses to the D-score. 

The `dscore` package currently support three keys: `dutch`, 
`gcdg` and `gsed`. Although there is much overlap, different keys 
cover different instrument. The table below displays the number of items 
per instrument under each of the three keys. If the entry is blank, 
the key does not cover the instrument.

For some instruments, e.g., for `cro` only one choice is possible (only 
key `gsed`). For `gri`, we may choose between `gcdg` and `gsed`. 
The actual choice will depend on the goals of your analysis. If you 
want to compare to other D-scores calculated under 
key `gcdg`, or reproduce an analysis made under key, then pick `gcdg`. 
If that is not the case, then
`gsed` is the better choice since it account for a wider variety of 
comparisons to other instruments. The differences between the D-scores
calculated under different keys are small, but these are not identical.
If you don't specify `key`, the `dscore()` function will use `key = "gsed"` 
for maximum instrument coverage.

The `dscore` package will recognize `r length(get_itemnames())` item 
names. Call the `get_itemnames()` function with any arguments to see
them all. If you want to see all names of a particular 
instrument and a particul domain, then use 

```{r}
# find all item name in mdt (MDAT), domain gm (gross motor)
items <- get_itemnames(instrument = "mdt", domain = "gm")
items
```

The labels for this set of items can be found by 

```{r}
head(get_labels(items))
```

Alternatively, use `decompose_itemnames()` to break down the item names 
into its components.

Using these functions, it should be relatively easy to map and rename
your item names into the GSED-convention.

## Resources

### Books and reports

1. [Introduction into the D-score](https://stefvanbuuren.name/dbook1/)
2. [Inventory of 147 instruments for measuring early child development](http://documents.worldbank.org/curated/en/384681513101293811/A-toolkit-for-measuring-early-childhood-development-in-low-and-middle-income-countries): @fernald2017

### Keys

1. Project with `dutch` key, 0-2 years: @vanbuuren2014
3. Project with `gcdg` key: @weber2019
4. Project with `gsed` key: @gsedteam2019

### Methodology

1. Interval scale: @jacobusse2006
2. Adaptive testing: @jacobusse2007 

### Literature


