---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# dscore: Measuring child development by the D-score

The D-score is a numerical score that measures generic development in children 0-4 years. The D-score is an attempt to analyze and predict development of children using tools developed for numerical measures, like height and weight.

## Installation

If you have been marked as a collaborator on GitHub, generated a personal access token (PAT) on <https://github.com/settings/tokens>, and stored your PAT by `Sys.setenv(GITHUB_PAT = "51he..")`, then install the `brokenstick` package from GitHub as follows: 

```{r eval = FALSE}
install.packages("devtools")
devtools::install_github(repo = "stefvanbuuren/dscore", oath = Sys.getenv("GITHUB_PAT"))
```


## Overview

The `dscore` package contains functions for

- Fitting and evaluating the Rasch model to data,
- Calculating individual trajectories of the D-score,
- Transforming the D-scores into Z-scores, and back.

## Main functions

The main functions in the `dscore` package are:

Function name        | Description
---------------------|---------------------------------
`rasch()`            | Estimate difficulty of items
`anchor_dscore()`    | Transform difficulties to fixed anchors
`dscore()`           | Estimate D-scores of children
`daz()`              | Transform to age-adjusted standardized D-score
`zad()`              | Inverse of `daz()`
