---
title: "Calculate D-score and DAZ"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calculate D-score and DAZ}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Overview

This vignettes shows how to estimate the D-score and the D-score SDS, a.k.a. DAZ in an excerpt from the POPS data. This vignettes covers some typical actions needed when estimating D-scores:

1. Rename item names in source data to item names used in itembank
2. Reorganize the source data into a long matrix
3. Calculate D-score and DAZ
4. Combine D-score and DAZ with source data

## Rename item names

The `dscore` package has built-in example data from the POPS study, called `popsdemo`. The data set is of class `tbl_df` from the `dplyr` package. 

```{r rename}
library("dscore")

popsdemo
class(popsdemo)
nrow(popsdemo)
```

The are 25 children and 4 time points. 

```{r children}
# 25 children, 4 time points per child
length(unique(popsdemo$patid))
```

The item scores that form the test are located in columns 9-65.
```{r}
test <- 9:65
```

These names of the columns need to be matched against one of the lexicons in the item bank. The built-in lexicons are:

```{r}
names(itembank)[1:6]
```

We first need to find out a proper lexicon for the data. For the POPS data, the closest lexicon is 
\code{lex.dutch1983}. Let us check the variable names in POPS with the item labels in the item bank.

```{r}
itemset <- !is.na(itembank$lex.dutch1983)
cbind(names(popsdemo)[test], itembank[itemset, c("lex.dutch1983", "labelEN", "tau")])
```

In this case, we are lucky that all item names from the source data and the item bank match up exactly. 
In general, we will need to map carefully the names in the dataset to the names in the item bank. For POPS, 
we may take out the relevant parts of the item bank as

```{r}
ib <- itembank[itemset,c("lex.dutch1983", "lex.GHAP", "labelEN", "tau")]
head(ib, 3)
```

From here on, we will work in the GHAP lexicon. Renaming the source data is now done by

```{r}
names(popsdemo)[test] <- as.character(ib$lex.GHAP)
```

The source data has now names that are recognized in the itembank. To check this, find the difficulties for each item by the `gettau()` function:

```{r}
gettau(names(popsdemo)[test])
```

## Reorganize the data into a long matrix

The `dscore()` function takes vectors of item scores, item names and ages. Rearringing the data makes it easy to extract the relevant vectors. We need to create a data set with the following variables: `patid`, `moment`, `age`, `daycor`, `item` and `score`, and select only the rows where we have an observed score.

```{r}
library("tidyr")
library("dplyr")
data <- popsdemo %>% 
  select(patid, moment, age, daycor, GSFIXEYE:GSKIK) %>%
  gather(items, scores, GSFIXEYE:GSKIK, na.rm = TRUE) %>%
  mutate(scores = 1 - scores) %>% 
  arrange(patid, moment)
data
```

There are `nrow(data)` records with a nonmissing item score. Note also that the item scores have been reversed, as POPS uses a zero for a PASS, and a one for a FAIL.

## Calculate D-score and DAZ

For illustration, let us first calculate the D-score of the first child. There are 75 scores for this child, spread over four time points. This is a preterm child, so we correct calener age for gestational age as in `daycor`:

```{r}
child1 <- filter(data, patid == 1)

scores <- child1$scores
items <- as.character(child1$items)
ages <- round(child1$daycor/365.25, 4)

# calculate dscore and daz for each time point for given child
(d <- dscore(scores, items, ages))
daz(d)
```

If desired, one may also back-calculate the D-score from the standard deviation score by 
```{r}
zad(daz(d))
```

If we specify the child identifier \code{patid} as a by-group variable, 
we may calculate the D-score and DAZ for all children by

```{r}
# use age corrected for gestational age
data <- data.frame(data)
data$ages <- round(data$daycor/365.25, 4)

# calculate D-score and DAZ
ds <- split(data, data$patid)
dl <- parallel::mclapply(ds, FUN = dscore)
dazl <- lapply(dl, FUN = daz)
df <- data.frame(
  patid = rep(as.numeric(names(dl)), times = unlist(lapply(dl, length))),
  ages = as.numeric(unlist(lapply(dl, names))),
  dscore = as.numeric(unlist(dl)),
  daz = as.numeric(unlist(dazl)))
head(df)
```

## Combine D-score and DAZ with source data

Finally, in order to do further analyses, we need to put the estimated D-score and DAZ back into the source data.

```{r}
# merge dscore and daz into popsdemo data
popsdemo$ages <- round(popsdemo$daycor/365.25, 4)
popsdemo <- merge(popsdemo, df, all.x = TRUE)
head(select(popsdemo, patid, moment, ages, dscore, daz))
```

